---
title: "What's in a Name?"
author: "Johanna Jantzen"
date: "September 30, 2020"
output: 
  pdf_document:
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
---
For instructors: If your students do not have internet access, R objects are available which contain the information retrieved from the online databases. Have your students read the files in (see Guide for Instructors for how to read R objects (.rds) into R) and use those objects instead of downloading the data from the online database. 

## Start of module

The code presented in this document is available in the R script entitled Taxonomy.R which can be edited and run within this R project. 

Before doing any work in R, you typically need to load the libraries with the functions that you will need. These libraries have previously been installed for this R project. Let's load them that now by selecting those rows and clicking "Run" or pressing *Ctrl Enter* (Windows) or *Cmd Return* (Mac). The warnings that you see in red are expected and can be ignored. In this pdf document, you can identify lines of code that were run because they start with the > character, while the resulting output is presented on lines that start with two hash marks *##*. In the pdf, lines of code which contain comments will therefore start with the symbols *>#*. Note that this is true only for the pdf, while in your R script (eg, Taxonomy.R) lines of code do not have any symbols at the start, while comments may start with any number of *#* symbols. 

```{r load libraries}

library(rotl)
library(dplyr)
library(ape)
library(taxize)

```

The Open Tree of Life project, which can be accessed at <https://opentreeoflife.github.io/>, has the goal of synthesizing a phylogeny for all of life, representing the evolutionary relationships between these organisms. A phylogeny that represents all known relationships between organisms, from plants to animals to microbes, has been made available for public use. If you are completing the entire course, we will start to investigate these relationships, focusing first on the names of taxa that we are studying in this "What's in a name?" module. 

Taxonomy is the system used to classify organisms into groups while Nomenclature is the system used to name these groups. There are several taxonomic databases that contain lists of species names and higher order taxonomies. This module will explore the names and diveristy of organisms in four **clades**: plants, insects, birds, and amphibians. Think about which of these four clades you would want to investigate further if you are continuing on to the phylogenetic and trait data modules in this course. 

##Import species lists

First, let's get a list of species to study. You can get species lists from any number of online sources (see Admin/Data_Attribution/Data_citations.txt for the sources of these species lists) or you can generate your own species lists for organisms you are interested in studing. We have selected these groups of organisms to generate five lists of taxa:

1)  Four conifer **genera** (plural of genus) that can be found around the world (Plants)

2)  The maple genus (*Acer*), including only the species found in Canada (Plants)

3)  Ant species occurring in Alberta, Canada (Insects)

4)  The nuthatch genus (*Sitta*) across its entire distribution (Birds)

5)  Frog species occurring in BC, Canada (Amphibians)

For each group of organisms (and therefore species lists), we are also including an **outgroup** species for downstream phylogenetic analysis (see "Module 2: Phylogenetics" for more information on outgroups).

We can generate R objects containing the lists of species names or genus names that we are interested in. For most of the groups, we are just typing the names into lists using the **function** *c()*. For the ant clade, we are reading in a **.txt** file which contains the list of species names. You can also open this "Ant_names.txt" file by clicking on the file name in the "Files" corner of RStudio in the folder "Data".

For birds, we'll look at the genus *Sitta* or nuthatches. We also include one species of wren (*Troglodytes*) as an outgroup for phylogenetic analysis. 

```{r list of birds, prompt = TRUE, tidy = TRUE}
birds <- c("Sitta", "Troglodytes aedon")
```

For insects, we're using a list of ant species from Alberta, Canada. We have an extra step of converting our object into a **vector** (list) instead of a **dataframe**. Note that this is not a **csv** file and we are using the **read.delim** function rather than the **read.csv** function to import this file
```{r list of ants, prompt = TRUE, tidy = TRUE}
ants_df <- read.delim("../Data/Ant_names.txt", header = FALSE, stringsAsFactors = FALSE)
str(ants_df)
ants <- as.vector(ants_df$V1)
str(ants)
```

For plants, we have two lists: a list of maple species from Canada with horse chestnut as an outgroup, and a list of conifer genera with *Ginkgo* as an outgroup.
```{r lists of plants, prompt = TRUE, tidy = TRUE}
maples <- c("Acer circinatum", "Acer glabrum", "Acer macrophyllum", "Acer negundo", "Acer nigrum", "Acer pensylvanicum", "Acer rubrum", "Acer saccharum", "Acer saccharinum", "Acer spicatum", "Aesculus glabra")
conifers <- c("Abies", "Thuja", "Picea", "Tsuga", "Ginkgo")
```

For amphibians, we have a list of frog species from BC, Canada with a salamander as an outgroup.
```{r list of frogs, prompt = TRUE, tidy = TRUE}
frogs <- c("Anaxyrus boreas", "Spea intermontanus", "Pseudacris maculata", "Hyla regilla", "Rana aurora", "Ascaphus truei", "Rana luteiventris", "Rana pretiosa", "Lithobates catesbeiana", "Lithobates clamitans", "Lithobates pipiens", "Lithobates sylvaticus", "Taricha granulosa")

```

## Assessment

1) To check that you have loaded the text file containing the ant species names properly and were able to run the lines of code above, how many species of ants were in the file? 

##Resolving names with TNRS

When dealing with species names, it's imporant to consider that there may be errors or inconsistencies in the names that you are using. The names in our lists could be old names or **synonyms** which may have been the result of taxonomic changes (such as moving a species from one genus to another) or could contain spelling or **orthographic** errors. Before we proceed with any data collection or analyses using these names, let's first resolve any of these "issues" using the Taxonomic Name Resolution Service (TNRS). 

We are using the **objects** generated by the code above with the **function** *tnrs_match_names* which does what it says: matches our species and genus names with names in the TNRS database. We assign the output from this **function** to a new **object**. It is good to use informative names for your **objects** so that you know what they contain. In this case, one new **object** contains the *birds* names that have been resolved so we use the name *birds_resolved*. You can name objects whatever you like (within reason) but short(ish) and concise names are best. Names also need to be one **string** so in place of spaces, you should use "_", "-", or "." to separate words, or you can use camel case like this: *birdsResolved*. 


``` {r taxonomy resolution, prompt = TRUE, tidy = TRUE}
#Match names using the taxonomic name resolution service
#These lines of code will not produce any output to the screen
birds_resolved <- tnrs_match_names(birds)
ants_resolved <- tnrs_match_names(ants)
maples_resolved <- tnrs_match_names(maples)
conifers_resolved <- tnrs_match_names(conifers)
frogs_resolved <- tnrs_match_names(frogs)

```

##Exploring the dataframes of resolved names

Now that we have checked our names against the TNRS, we can see if there have been any name changes or errors. Let's look at the dataframes that were output by the *tnrs_match_names* **function**. There are several ways to look at our data. 

We can look at the entire object by just **running** the object. This works best when the objects are small (that is, they do not have too many rows or columns). We can also inspect just the first few rows using the **function** *head()* or we can select specific rows or columns by either subsetting using a number within square brackets *[ ]* or the column name using *$*.  

Remember that we can use a variety of **functions** to get more information about the dimensions and structure of **objects**:

* *nrow()* - get the number of rows of a dataframe/table

* *ncol()* - get the number of columns of a dataframe/table

* *rownames()* - get the names of the rows of a dataframe/table

* *colnames()* - get the names of the columns of a dataframe/table

* *dim()* - get the dimensions (height and width) of a dataframe/table

* *str()* - get the structure of an object

* *class()* - get the class of an object

* *summary()* - get a summary of an object

* *length()* - get the length of an object (eg a list)

``` {r inspecting objects,  prompt = TRUE, tidy = TRUE}

#Inspect entire bird object
birds_resolved
nrow(maples_resolved)

#Get the first few rows of the ants object
head(ants_resolved)

#Get the first and third rows of the maples object
maples_resolved[c(1,3),]

#Get the first to third rows and the first to fifth columns of the conifers object
conifers_resolved[1:3, 1:5]

#Get the "unique_name" and "is_synonym" columns of the frogs object
frogs_resolved$unique_name
frogs_resolved$is_synonym

#To identify other columns which may contain helpful information, get a list of column names
colnames(frogs_resolved)

#For those datasets which now include the correct species names, we can make our new species name lists
frog_species_list <- frogs_resolved$unique_name
maple_species_list <- maples_resolved$unique_name
ant_species_list <- ants_resolved$unique_name



```

Let's see if you can get some more information about of these objects. 

## Assessment

2) How many rows does the *maples_resolved* object have? What function(s) can you use to calculate that?
3) How many columns does the *birds_resolved* object have? What function(s) can you use to calculate that?
4) How many taxa (from all five datasets) have synonyms? Which column contains this information?
5) Which dataset(s) (**object(s)**) contain(s) taxa that have no taxonomic problems, ambiguities or name changes/synonyms (at the taxonomic level studied)? Explain.
6) Give an example of an **orthographic** or spelling error from one of the datasets (**objects**) and one example of a synonym due to the movement from a species from one genus to another.
7) What might you conclude about the taxonomy of the ant species? Do you have confidence in the names that were matched?
8) Are you able to determine which of the five datasets contains the most species? Why or why not?


##Retrieving names for lower taxonomic levels

You may have noticed that for some of the datasets, we have  lists of **species binomials** composed of **genus** names and **specific epithets**. For others, however, we only have the **genus** name and no information on the **species** within that **genus**. If we want to get a list of all the species that are in a genus, we can use one of these **functions**: *downstream()* or *children()* from the **package** *taxize*. 

*downstream()* can obtain names for different taxonomic levels below the one you **query**. For example, if you are looking for a list of **genera** within the plant **family** "Sapindaceae", the **function** would look like: 
*downstream("Sapindaceae", downto = "genus", db = "itis")*. 
Alternatively, you could get the list of **species** within this family by specifying:
*downstream("Sapindaceae", downto = "species", db = "itis")*. 

*children()* is slightly different in that this **function** only retrieves the names at the taxonomic level immediately below the one queried. So if you input a list of **genera**, you will receive the **species** names, while if you input a list of **families**, you would receive a list of **genera**. 

These **functions** can access several different taxonomic databases. In this module, we will use the Integrated Taxonomic Information System (ITIS) which can also be accessed through <https://www.itis.gov/>. 
The first element in these **functions** can either be a taxonomic name in quotes (eg "Sapindaceae") or an R **object** that contains a list of taxon names which should not be put in quotation marks (eg conifers). You can specify the taxonomic level you are searching using the *downto* field for the *downstream()* function. To specify which database you want to query, use the *db* field (we are using the ITIS database).
```{r get lower level taxonomic names,  prompt = TRUE, tidy = TRUE}

#Examples of *downstream()* and *children()* functions

#To find more information on specific functions, you can use the following command 
#(see the help tab on the bottom right of the screen)
#?downstream

#Here, we have specified to get names down to the species level
sapindaceae_species <- downstream("Sapindaceae", downto = "species", db = "itis")

#The children function only retrieves the level immediately below, so here we are retrieving the genus names
sapindaceae_genera <- children("Sapindaceae", db = "itis")

```

The next section has some code that is a bit more complicated but uses the functions we tested with Sapindaceae above. As you run these lines, see if you can figure out how each line works. 

Now we will get species names for those datasets without **specific epithets** (ie birds, conifers). Note that for our bird dataset, we don't want to get species name information for the outgroup because it is already a **species binomial** not a **genus** name. 

```{r print birds,  prompt = TRUE, tidy = TRUE}

birds
```

Although we could just type the name *Sitta* in as our query, we will edit our list by omitting the last entry (the **outgroup**) from the list. While it would be easy to do this manually for our very short list of taxa, if we had a longer list it wouldn't be practical. We'll break this down into its component steps.
Note: you do not need to change any of the code for this section and don't worry if you do not understand the code; this may be a first introduction for you so just follow along with this demonstration of how to manipulate objects in R so you can see what code can do. 

```{r Get species names for birds and conifers,  prompt = TRUE, tidy = TRUE}
#First, we want to know how long the list is
length(birds)

#We want to get rid of the last entry (the Nth entry for a list of length N) for this query so we use the '-' symbol to indicate we want to remove this entry, using the [ ] we had used earlier to subset our data.
#As seen above, length() is a function to get the length of a list
birds[-(length(birds))]

#Because length(birds) = 2, the code can also be written as 
birds[-2]

#We now have the version of the list we want for the query so we can insert it into our function
species_birds <- downstream(birds[-(length(birds))], downto = "species", db = "itis")

#We can compare the other function for getting species names for same bird dataset. Are the results the same?
species_birds2 <- children(birds[-(length(birds))], db = "itis")
species_birds2$Sitta

#If we want to create our final list of bird species including our outgroup, we can now add our last entry from the original taxon list back (omitting the "-" now becasue we want to include only that entry instead of exlcuding it)
bird_species_list <- c(species_birds$Sitta$taxonname, birds[(length(birds))])

#We can now get species names for conifers - all the names in the list are genus names so we can query the entire list
species_conifers <- downstream(conifers, downto = "species", db = "itis")

```
Notice the difference in the way that the *downstream()* function is written in the first example compared to with the conifer dataset. 

sapindaceae_species <- downstream("Sapindaceae", downto = "species", db = "itis")

species_conifers <- downstream(conifers, downto = "species", db = "itis")

In the first example, the query was put in quotation marks ("Sapindaceae") while in the second, the query was not (conifers). Why do you think this is? 

**Spoiler**: the difference in syntax has to do with whether the query is to be searched verbatim (in quotation marks) or if it represents an object containing the text to be searched. For the example using the family Sapindaceae, the name we wanted to find was the text that we queried in the function: Sapindaceae. The *downstream()* function then searched the ITIS database (specified using the term *db = "itis"*) for the exact phrase *"Sapindaceae"* and returned the resulting list of species (as specified using the  term *downto = "species"*). However, when searching for the genera represented by the object *conifers*, we do not want the *downstream()* function to search the database for the text *"conifers"*, which is what would happen if we included those quotation marks, because *"conifers"* is not an official taxonomic name. Instead, we specify the object *conifers* so that the function will run for all the individual taxa included in the object (that is, the five genera listed in the **conifers** object). So to summarize, if you include quotation marks around text, functions (including *downstream()* and *children()*) will typically treat that text as the query itself, while if you omit quotation marks, the function will recognize the input as an object which stores one or more queries.

Now you have a **list** of **dataframes** for the conifers with a different **dataframe** for each **genus**, but you want the names as a single **list**. We will now do some data manipulation that R makes simpler and more reproducible than cutting and pasting in Excel. We can retrieve the list of species within each genus using the *$* symbol which retrieves a specific column. For example, within the *species_conifers* **list**, we can specify the *Abies* **dataframe** and the *taxonname* **column** which includes the species names. 
```{r Reformat species names for conifers,  prompt = TRUE, tidy = TRUE}
species_conifers

species_conifers$Abies

species_conifers$Abies$taxonname

#We will take these five lists of names and join them into one list using the c() function
conifer_species_list <- c(species_conifers$Abies$taxonname, species_conifers$Thuja$taxonname, species_conifers$Picea$taxonname, species_conifers$Tsuga$taxonname, species_conifers$Ginkgo$taxonname)


```
We now have five lists of species for our five groups of organisms:

* conifer_species_list

* bird_species_list

* frog_species_list

* maple_species_list 

* ant_species_list 

Now it's time for you to explore these databases further for your assessment. 

Tip: Retrieving the data for some of the functions can take a little while so while you are waiting for one function to run, you can think about and write the code for the next question.


## Assessment

9) How many **species** (including **hybrid species** which are indicated by "X") of *Abies* were retrieved? How many of *Picea*?
10) Pick a **genus** that you're fond of and retrieve the names of **species** within that **genus** using one of the two **functions** illustrated in this module. *Requires internet* - if no internet, write the code you would use to do this
11) Retrieve the taxonomic level below species for the frog dataset. What do you find? Do the same for the conifer dataset. What about for the conifer dataset? (Hint: start with the **objects** containing the lists of **species** for frogs and for conifers. Copy an example of the code given earlier and edit it in a new line. Remember that correct punctuation is important.)
12) Now, which of our five datasets has the most species? How did you calculate that?

## Saving lists of species for future analysis

Now that we have our lists of species, we may want to save those names, or we may want to save the dataframes of resolved names that contain the OTT IDs so that we can query the Open Tree of Life to retrieve a phylogeny for our species. We will do both here to prepare ourselves for the next module in which we retrieve and visualize phylogenies for our organisms of interest. 

The **objects** that are created in R are not saved unless we tell the program to save them. In R Projects, sometimes the **environment** can be saved, so objects that were read into R can be used the next time you open the Project. However, sometimes the results you produce need to be able to be shared with others which requires you to save the **object** to a file. 

Here, we will save our lists of species as **.txt** files and our dataframes containing the OTT IDs as **.csv** files. We need to specify a **path** to where the files will be saved. Because we are working in an R Project, the **path** starts where our script is, so first we move up one level to the "Module_1" folder (*../*), and then specify the **Output** folder (*../Output/*) and the filename of our choice (*../Output/xxxx_species_names.txt*). Each time you enter a new folder or move back a level in the folder structure (*../*) a new section of the path needs to be added (separated by */*). We specify that we want our output files of species names to be **tab delimited** using the **function** *write.table()* and the term *sep = "\ t"*. This means that each column would be separated by a tab in the file. We specify that we want our dataframes including the OTT IDs to be saved as **comma delimited files** using the **function** *write.csv()*. The endings of the file names typically indicate what type of file it is (eg **.csv** versus **.txt**). 

``` {R saving objects,  prompt = TRUE, tidy = TRUE}

#write the objects to files - saving these files for future reference
write.table(conifer_species_list, "../Output/conifer_species_names.txt", sep = "\t", col.names = FALSE, row.names = FALSE)
write.table(bird_species_list, "../Output/bird_species_names.txt", sep = "\t", col.names = FALSE, row.names = FALSE)
write.table(frog_species_list, "../Output/frog_species_names.txt", sep = "\t", col.names = FALSE, row.names = FALSE)
write.table(maple_species_list, "../Output/maple_species_names.txt", sep = "\t", col.names = FALSE, row.names = FALSE)
write.table(ant_species_list, "../Output/ant_species_names.txt", sep = "\t", col.names = FALSE, row.names = FALSE)

write.csv(birds_resolved, "../Output/birds_OTT_IDs.csv", row.names = FALSE)
write.csv(conifers_resolved, "../Output/conifers_OTT_IDs.csv", row.names = FALSE)
write.csv(frogs_resolved, "../Output/frogs_OTT_IDs.csv", row.names = FALSE)
write.csv(maples_resolved, "../Output/maples_OTT_IDs.csv", row.names = FALSE)
write.csv(ants_resolved, "../Output/ants_OTT_IDs.csv", row.names = FALSE)

```

Congratulations! You have resolved taxonomic issues for several groups of organisms, retrieved OTT IDs for these taxa, obtained a list of species names for higher order taxa, and saved your species names and dataframes as files for future analysis. 
